### 2021-08-20 bare_jrnl.tex
Initial Overleaf Import

\documentclass[journal]{IEEEtran}

\usepackage[pdftex]{graphicx}
\usepackage{amsmath}
\usepackage{algorithmic}
\usepackage{cleveref}
\usepackage{url}

\begin{document}

\title{TBC: In-situ visualization with TACC Galaxy in Material Point Method}
% author names and IEEE memberships%

\author{Greg Abram, %~\IEEEmembership{Fellow,~OSA,}
        Andrew Solis,
        and~Krishna Kumar% <-this % stops a space
\thanks{Department of Civil, Architectural and Environmental Engineering, University of Texas at Austin, TX, 78712 USA e-mail: (krishnak@utexas.edu).}% <-this % stops a space
\thanks{Texas Advanced Computing Center, University of Texas at Austin, TX,.}% <-this % stops a space
\thanks{Manuscript received September 6th, 2021.}}

% The paper headers
\markboth{IEEE Computing in Science and Engineering (CiSE), Special Issue 2021}%
{}

%\IEEEpubid{0000--0000/00\$00.00~\copyright~2015 IEEE}
% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
The abstract goes here.
\end{abstract}

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
In-situ viz, MPM, TACC-Galaxy.
\end{IEEEkeywords}

\IEEEpeerreviewmaketitle

\section{Overview and motivation}

\IEEEPARstart{L}{andslide} runouts are regional-scale events that can bury whole towns (e.g., 2018 Southern California mudflows that followed a series of wildfires~\cite{lukashov2019post}) or devastate entire regions (e.g., 2016 Kaikoura, New Zealand earthquake recorded 100,000 landslides within a 12,000 sq. km area~\cite{dellow2017landslides}). The frequency of these regional-scale landslides is increasing with devastating earthquakes, extreme precipitation, and wildfires caused by climate change. Even in these risk-prone communities, where the residents are aware of the threat posed by the landslides and in some cases experienced significant modification to the landscape or their behavior, the residents repeatedly show indifference to the threats, consequently fail to act. The human capacity to deny danger endangers lives and property~\cite{kim2020public}. In 2020, twenty-two extreme events cost the United States \$95 billion in damages. The underlying problem is the uncertainty of the actual event and a lack of understanding of its potential impacts. How can the reality of potential risk be communicated to a diverse group of actors, who must understand and support a complex set of actions to reduce that risk for the whole community?

Visualization is the key to communicating scientific results effectively to engineering decision-makers and the public~\cite{mayer2005cognitive}. Given limited cognitive capacity, visualizing the complex information of threats as images enhances the brain's capacity to perceive opportunities and make decisions. Visual representation of complex information reduces the cognitive load on human information processing and increases human "absorptive capacity" for problem-solving~\cite{kahneman2012human}. However, these visualizations must be physically sound to be perceived as realistic and accepted as valid.

Despite the landslide risks, most regional-scale landslide hazard analyses do not consider downstream impacts and the aerial extent of debris-flow runouts~\cite{USGS2017debris}. Traditional numerical methods such as the Finite Element Method (FEM) primarily focuses on the onset of failure but provides limited information on the post-failure runout mechanisms due to mesh distortions at large  deformations~\cite{soga2016trends}. Modeling the impacts of the potential runout of large landslides is possible with new tools such as Material Point Method (MPM)~\cite{Bardenhagen2000,Bardenhagen2004,Sulsky1994,Sulsky1995}. MPM is a mesh-free method that discretizes the domain as a collection of \textit{material points} moving on a background grid, and their deformations are determined by Newton's laws of motion. In this study, we employ the CB-Geo MPM code (\url{https://github.com/cb-geo/mpm}), and a typical MPM computation cycle is shown in~\cref{fig:mpm}. For more information on MPM and the code implementation, refer to~\cite{kumar2019scalable}. MPM results, which can then be used to translate the results of these analyses into effective visualizations to inform decision-makers of potential hazards.

Visualization is also context and user-specific, i.e., different users may be interested in different aspects of the experimental or simulation data sets. Hence, the same data set may be represented in several forms – perhaps at varying levels of detail, emphasizing or deemphasizing different regions and features, and employing different visualization techniques to best present the information. Data retrieval and visualization in High-Performance Computing applications have long been a bottleneck. Current techniques for visualization petascale simulation output involve writing a temporal slice of a subset of the data to disk. In practice, this usually means that the data are stored only at specific time-steps, often at a much coarser resolution than the original data, which leads to a significant portion of information being disregarded and potentially lost. A large-scale simulation running for thousands of time steps across compute nodes sampled at a predefined frequency of only every 100 - 500-time steps results in several terabytes of data. Nevertheless, the scale of such discrete geometry visualization will push commercial rendering engines like Mantra and Arnold to the limit. The amount of data would be several orders of magnitude higher at exascale levels, making the simulation spend most of the supercomputing time doing I/O.

In-situ visualization is a rendering technique for visualizing the simulation data in real-time without requiring additional storage resources. Running the visualization and simulation in tandem avoids the bottleneck of data transfer. Furthermore, this approach allows for monitoring and interaction while the simulation is running, which enables scientists to modify simulation parameters and explore the effect on the phenomena in real-time. In-situ visualization is a promising solution, and libraries such as ParaView Catalyst and SENSEI offer the ability to couple existing data models (such as VTK) to query regions of interest and to render simulations. Ray tracing engines such as TACC's Gravit and Galaxy offer distributed asynchronous in-situ visualization capabilities for petascale simulations.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=\linewidth]{figs/mpm.pdf}
    \caption{Illustration of the MPM algorithm (1) A representation of material points overlaid on a computational grid. Arrows represent material point state vectors (mass, volume, velocity, etc.) being projected to the nodes of the computational grid. (2) The equations of motion are solved onto the nodes, resulting in updated nodal velocities and positions. (3) The updated nodal kinematics are interpolated back to the material points. (4) The state of the material points is updated, and the computational grid is reset.}
    \label{fig:mpm}
\end{figure}
\begin{itemize}
  \item why simulating earthquakes is important
  \item why such sim is hard
  \item brief description of MPM and benefits, cite to marker paper
  \item why in situ vis is needed / benefits
  \item Galaxy + MPM = awesomeness
\end{itemize}

\section{Related work}
  - brief summary other in situ, cite Hank term paper, maybe others

\section{Methodology}
\begin{itemize}
  \item tie into existing VTK-write path
  \item Galaxy expand to handle unstructured mesh
  \item tracking particles across domain boundaries
\end{itemize}

\section{Results}
  - show some pretty pictures
  - provide some runtime stats

\section{Conclusion}
- Future work

\appendices

% use section* for acknowledgment
\section*{Acknowledgment}
The authors would like to thank...

% Can use something like this to put references on a page
% by themselves when using endfloat and the captionsoff option.
\ifCLASSOPTIONcaptionsoff
  \newpage
\fi

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{IEEEabrv,references.bib}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
%\bibliography{references.bib}

% biography section
% 
% If you have an EPS/PDF photo (graphicx package needed) extra braces are
% needed around the contents of the optional argument to biography to prevent
% the LaTeX parser from getting confused when it sees the complicated
% \includegraphics command within an optional argument. (You could create
% your own custom macro containing the \includegraphics command to make things
% simpler here.)
%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
% or if you just want to reserve a space for a photo:

% \begin{IEEEbiography}{Michael Shell}
% Biography text here.
% \end{IEEEbiography}

% if you will not have a photo at all:
% \begin{IEEEbiographynophoto}{John Doe}
% Biography text here.
% \end{IEEEbiographynophoto}

%\vfill

% Can be used to pull up biographies so that the bottom of the last one
% is flush with the other column.
%\enlargethispage{-5in}
\end{document}

### 2021-08-20 bare_jrnl.tex
:hammer: Revise intro and motivation

@@
 
-\title{TBC: In-situ visualization with TACC Galaxy in Material Point Method}
-% author names and IEEE memberships%
-
+\title{In-situ visualization of natural hazards with TACC Galaxy and Material Point Method}
 \author{Greg Abram, %~\IEEEmembership{Fellow,~OSA,}
@@
 \thanks{Texas Advanced Computing Center, University of Texas at Austin, TX,.}% <-this % stops a space
-\thanks{Manuscript received September 6th, 2021.}}
+\thanks{Manuscript received September 3rd, 2021.}}
 
@@
 \section{Overview and motivation}
-
-\IEEEPARstart{L}{andslide} runouts are regional-scale events that can bury whole towns (e.g., 2018 Southern California mudflows that followed a series of wildfires~\cite{lukashov2019post}) or devastate entire regions (e.g., 2016 Kaikoura, New Zealand earthquake recorded 100,000 landslides within a 12,000 sq.
+% - why simulating earthquakes is important
+% - why such sim is hard
+% - brief description of MPM and benefits, cite to marker paper
+% - why in situ vis is needed / benefits
+% - Galaxy + MPM = awesomeness
+\IEEEPARstart{L}{andslide} runouts are regional-scale events that can bury whole towns (e.g., 2018 Southern California mudflows that followed a series of wildfires~\cite{lukashov2019post}, 2014 Oso landslide caused 43 fatalities in the US) or devastate entire regions (e.g., 2016 Kaikoura, New Zealand earthquake recorded 100,000 landslides within a 12,000 sq.
 km area~\cite{dellow2017landslides}).
 The frequency of these regional-scale landslides is increasing with devastating earthquakes, extreme precipitation, and wildfires caused by climate change.
-Even in these risk-prone communities, where the residents are aware of the threat posed by the landslides and in some cases experienced significant modification to the landscape or their behavior, the residents repeatedly show indifference to the threats, consequently fail to act.
+Even in these risk-prone communities, where the residents are aware of the threat posed by the landslides, the residents repeatedly show indifference to these threats and consequently fail to act.
 The human capacity to deny danger endangers lives and property~\cite{kim2020public}.
@@
 The underlying problem is the uncertainty of the actual event and a lack of understanding of its potential impacts.
-How can the reality of potential risk be communicated to a diverse group of actors, who must understand and support a complex set of actions to reduce that risk for the whole community?
+How to communicate the reality of potential risk to a diverse group of individuals, who must understand and support a complex set of actions that reduce the risk for the whole community?
 
 Visualization is the key to communicating scientific results effectively to engineering decision-makers and the public~\cite{mayer2005cognitive}.
-Given limited cognitive capacity, visualizing the complex information of threats as images enhances the brain's capacity to perceive opportunities and make decisions.
-Visual representation of complex information reduces the cognitive load on human information processing and increases human "absorptive capacity" for problem-solving~\cite{kahneman2012human}.
+Given the limited cognitive capacity in humans, visualizing the complex threats as images enhances the brain's capacity to perceive opportunities and make decisions.
+Visual representation of complex information reduces the cognitive load on human information processing and increases human ``absorptive capacity" for problem-solving~\cite{kahneman2012human}.
 However, these visualizations must be physically sound to be perceived as realistic and accepted as valid.
@@
 Modeling the impacts of the potential runout of large landslides is possible with new tools such as Material Point Method (MPM)~\cite{Bardenhagen2000,Bardenhagen2004,Sulsky1994,Sulsky1995}.
-MPM is a mesh-free method that discretizes the domain as a collection of \textit{material points} moving on a background grid, and their deformations are determined by Newton's laws of motion.
+MPM is a mesh-free method that discretizes the domain as a collection of \textit{material points} moving on a background grid, and Newton's laws of motion determine their deformations.
 In this study, we employ the CB-Geo MPM code (\url{https://github.com/cb-geo/mpm}), and a typical MPM computation cycle is shown in~\cref{fig:mpm}.
 For more information on MPM and the code implementation, refer to~\cite{kumar2019scalable}.
-MPM results, which can then be used to translate the results of these analyses into effective visualizations to inform decision-makers of potential hazards.
-
-Visualization is also context and user-specific, i.e., different users may be interested in different aspects of the experimental or simulation data sets.
-Hence, the same data set may be represented in several forms – perhaps at varying levels of detail, emphasizing or deemphasizing different regions and features, and employing different visualization techniques to best present the information.
-Data retrieval and visualization in High-Performance Computing applications have long been a bottleneck.
-Current techniques for visualization petascale simulation output involve writing a temporal slice of a subset of the data to disk.
-In practice, this usually means that the data are stored only at specific time-steps, often at a much coarser resolution than the original data, which leads to a significant portion of information being disregarded and potentially lost.
-A large-scale simulation running for thousands of time steps across compute nodes sampled at a predefined frequency of only every 100 - 500-time steps results in several terabytes of data.
-Nevertheless, the scale of such discrete geometry visualization will push commercial rendering engines like Mantra and Arnold to the limit.
-The amount of data would be several orders of magnitude higher at exascale levels, making the simulation spend most of the supercomputing time doing I/O.
-
-In-situ visualization is a rendering technique for visualizing the simulation data in real-time without requiring additional storage resources.
-Running the visualization and simulation in tandem avoids the bottleneck of data transfer.
-Furthermore, this approach allows for monitoring and interaction while the simulation is running, which enables scientists to modify simulation parameters and explore the effect on the phenomena in real-time.
-In-situ visualization is a promising solution, and libraries such as ParaView Catalyst and SENSEI offer the ability to couple existing data models (such as VTK) to query regions of interest and to render simulations.
-Ray tracing engines such as TACC's Gravit and Galaxy offer distributed asynchronous in-situ visualization capabilities for petascale simulations.
+MPM results of the landslide runout can be translated into compelling visualizations to inform decision-makers of potential hazards.
 
@@
     \caption{Illustration of the MPM algorithm (1) A representation of material points overlaid on a computational grid.
-Arrows represent material point state vectors (mass, volume, velocity, etc.) being projected to the nodes of the computational grid.
+Arrows represent material point state vectors (e.g., mass, volume, and velocity) is projected to the nodes of the computational grid.
 (2) The equations of motion are solved onto the nodes, resulting in updated nodal velocities and positions.
-(3) The updated nodal kinematics are interpolated back to the material points.
+(3) The updated nodal kinematics is interpolated back to the material points.
 (4) The state of the material points is updated, and the computational grid is reset.}
@@
 \end{figure}
-\begin{itemize}
-  \item why simulating earthquakes is important
-  \item why such sim is hard
-  \item brief description of MPM and benefits, cite to marker paper
-  \item why in situ vis is needed / benefits
-  \item Galaxy + MPM = awesomeness
-\end{itemize}
+
+A regional-scale landslide simulation of a kilometer-scale runout with MPM requires billions of material points and millions of cells in the background grid.
+Scaling applications to simulate a billion particles require exascale high-performance computing (HPC) architectures with efficient Message-Passing Interface (MPI).
+This study uses the CB-Geo MPM code, which exploits a hybrid MPI+X approach to achieve regional-scale simulation of landslides.
+In addition, as the landslide runout progresses, the workload distributed across the compute nodes becomes imbalanced.
+We adopt a distributed graph-partitioning technique to load balance both the background mesh and particle distribution dynamically.
+
+Visualization is also context and user-specific, i.e., different users may be interested in different aspects of the experimental or simulation data sets.
+Hence, the same data set may be represented in several forms – perhaps at varying levels of detail, emphasizing or deemphasizing different regions and features, and employing different visualization techniques to best present the information.
+Data retrieval and visualization in High-Performance Computing applications have long been a bottleneck.
+Current techniques for visualizing exascale simulations involve writing a temporal slice of a subset of the data to disk, often at a much coarser resolution than the original data, which leads to a significant portion of information being disregarded and potentially lost, followed by post-processing.
+In practice, this usually means that the data are stored only at specific time steps.
+An exascale simulation would output terabytes of data, resulting in a significant portion of the compute time spent on I/O~\cite{byna2020exahdf5}.
+Nevertheless, the scale of such discrete geometry visualization will push commercial rendering engines like Mantra and Arnold to the limit.
+The amount of data would be several orders of magnitude higher at exascale levels, making the simulation spend most of the compute time in I/O.
+
+In-situ visualization is a rendering technique for visualizing the simulation data in real-time without requiring additional storage resources~\cite{abram2018galaxy,atzori2021situ,kress2019comparing}.
+Running the visualization and simulation in tandem avoids data transfer bottleneck, enabling scientists to monitor and modify simulation parameters in real-time.
+In-situ visualization libraries such as ParaView Catalyst and SENSEI offer the ability to couple existing data models (such as VTK) to query regions of interest and to render simulations.
+However, scaling rendering engines to exascale simulations remains a challenge requiring billions of lightpaths to be rendered.
+Most rendering engines only sample a limited number of random light paths can be at each step.
+Thus, the visual quality of rendered images suffers from estimator variance, which appears as visually distracting noise~\cite{yang2021foveated}.
+Ray tracing engines such as TACC Galaxy~\cite{abram2018galaxy} offer distributed asynchronous in-situ visualization capabilities for exascale simulations.
+In this study, we couple the CB-Geo MPM code with distributed asynchronous ray tracing engine TACC Galaxy to realize the 2014 Oso landslide in the US with 50 million material points.
 

### 2021-09-06 bare_jrnl.tex
:pencil: First draft

@@
 
-\title{In-situ visualization of natural hazards with TACC Galaxy and Material Point Method}
+\title{In-situ visualization of natural hazards with Galaxy and Material Point Method}
 \author{Greg Abram, %~\IEEEmembership{Fellow,~OSA,}
         Andrew Solis,
+        Yong Liang,
         and~Krishna Kumar% <-this % stops a space
@@
 \thanks{Texas Advanced Computing Center, University of Texas at Austin, TX,.}% <-this % stops a space
+\thanks{University of California at Berkeley, CA,.}% <-this % stops a space
 \thanks{Manuscript received September 3rd, 2021.}}
@@
 \begin{abstract}
-The abstract goes here.
+Visualizing regional-scale landslides is the key to conveying the threat of natural hazards to stakeholders and policymakers.
+Traditional visualization techniques are restricted to post-processing a limited subset of simulation data and are not scalable to rendering exascale models with billions of particles.
+In-situ visualization is a technique of rendering simulation data in real-time, i.e., rendering visuals in tandem while the simulation is running.
+In this study, we develop a scalable N:M interface architecture to visualize regional-scale landslides.
+We demonstrate the scalability of the architecture by simulating the long runout of the 2014 Oso landslide using the Material Point Method coupled with the Galaxy ray tracing engine rendering 4.2 million material points as spheres.
+In-situ visualization has an amortized runtime increase of 2\% compared to non-visualized simulations.
+The developed approach can achieve in-situ visualization of regional-scale landslides with billions of particles with minimal impact on the simulation process.
 \end{abstract}
@@
 \begin{IEEEkeywords}
-In-situ viz, MPM, TACC-Galaxy.
+In-situ visualization, Material Point Method, TACC Galaxy, Ray tracing.
 \end{IEEEkeywords}
@@
 % - Galaxy + MPM = awesomeness
-\IEEEPARstart{L}{andslide} runouts are regional-scale events that can bury whole towns (e.g., 2018 Southern California mudflows that followed a series of wildfires~\cite{lukashov2019post}, 2014 Oso landslide caused 43 fatalities in the US) or devastate entire regions (e.g., 2016 Kaikoura, New Zealand earthquake recorded 100,000 landslides within a 12,000 sq.
-km area~\cite{dellow2017landslides}).
+\IEEEPARstart{L}{andslide} runouts are regional-scale events that can bury whole towns (e.g., 2018 Southern California mudflows that followed a series of wildfires, 2014 Oso landslide caused 43 fatalities in the US) or devastate entire regions (e.g., 2016 Kaikoura, New Zealand earthquake recorded 100,000 landslides within a 12,000 sq.
+km area).
 The frequency of these regional-scale landslides is increasing with devastating earthquakes, extreme precipitation, and wildfires caused by climate change.
@@
 
-Visualization is the key to communicating scientific results effectively to engineering decision-makers and the public~\cite{mayer2005cognitive}.
+Visualization is the key to communicating scientific results effectively to engineering decision-makers and the public.
 Given the limited cognitive capacity in humans, visualizing the complex threats as images enhances the brain's capacity to perceive opportunities and make decisions.
@@
 However, these visualizations must be physically sound to be perceived as realistic and accepted as valid.
+Visualization is also context and user-specific, i.e., different users may be interested in different aspects of the experimental or simulation data sets.
+Hence, the same data set may be represented in several forms – perhaps at varying levels of detail, emphasizing or deemphasizing different regions and features, and employing different visualization techniques to best present the information.
 
@@
 Traditional numerical methods such as the Finite Element Method (FEM) primarily focuses on the onset of failure but provides limited information on the post-failure runout mechanisms due to mesh distortions at large  deformations~\cite{soga2016trends}.
-Modeling the impacts of the potential runout of large landslides is possible with new tools such as Material Point Method (MPM)~\cite{Bardenhagen2000,Bardenhagen2004,Sulsky1994,Sulsky1995}.
+Modeling the impacts of the potential runout of large landslides is possible with new tools such as Material Point Method (MPM)~\cite{kumar2019scalable}.
 MPM is a mesh-free method that discretizes the domain as a collection of \textit{material points} moving on a background grid, and Newton's laws of motion determine their deformations.
@@
 
-Visualization is also context and user-specific, i.e., different users may be interested in different aspects of the experimental or simulation data sets.
-Hence, the same data set may be represented in several forms – perhaps at varying levels of detail, emphasizing or deemphasizing different regions and features, and employing different visualization techniques to best present the information.
 Data retrieval and visualization in High-Performance Computing applications have long been a bottleneck.
@@
 
-In-situ visualization is a rendering technique for visualizing the simulation data in real-time without requiring additional storage resources~\cite{abram2018galaxy,atzori2021situ,kress2019comparing}.
 Running the visualization and simulation in tandem avoids data transfer bottleneck, enabling scientists to monitor and modify simulation parameters in real-time.
@@
 However, scaling rendering engines to exascale simulations remains a challenge requiring billions of lightpaths to be rendered.
-Most rendering engines only sample a limited number of random light paths can be at each step.
+Most rendering engines only sample a limited number of random light paths at each step.
 Thus, the visual quality of rendered images suffers from estimator variance, which appears as visually distracting noise~\cite{yang2021foveated}.
-Ray tracing engines such as TACC Galaxy~\cite{abram2018galaxy} offer distributed asynchronous in-situ visualization capabilities for exascale simulations.
-In this study, we couple the CB-Geo MPM code with distributed asynchronous ray tracing engine TACC Galaxy to realize the 2014 Oso landslide in the US with 50 million material points.
+Ray tracing engines such as TACC Galaxy~\cite{abram2018galaxy} (\url{https://github.com/cb-geo/mpm}) offer distributed asynchronous in-situ visualization capabilities for exascale simulations.
+
+This study develops an in-situ visualization technique to simulate and render regional-scale landslides by coupling the CB-Geo MPM code for large-deformation modeling with distributed asynchronous ray tracing engine TACC Galaxy.
+In this study, we apply the in-situ visualization technique in MPM to simulate and visualize the disastrous 2014 Oso landslide, also known as the SR 530 landslide, which occurred in the northwest Washington state in the US, resulting in an 1100 m runout.
+
+\begin{figure}[!htbp]
+    \centering
+    \includegraphics[width=\linewidth]{figs/osofull.png}
+    \caption{Visualization of a timestep of the MPM simulation of the Oso landslide.
+Particles are represented as spheres with a radius approximating the original resolution.
+Color represents the distance of each particle from its original location.}
+    \label{fig:osofull}
+\end{figure}
+
+\Cref{fig:osofull} shows a visualization of the final runout of the MPM simulation of the Oso landslide.
+In this case, we use Galaxy to render the scene using spheres to represent the material points in the simulation, each with a fixed radius matching the original particle grid resolution and colored by the distance between each particle's current and original location.
+Galaxy's global lighting model is used to calculate shadows for additional depth perception.
 
 \section{Related work}
-  - brief summary other in situ, cite Hank term paper, maybe others
+\subsection{Visualization Systems}
+Systems for visualizing large-scale scientific data have been available for decades.
+High-performance graphics engines - add-ons to powerful workstations first made interactive post-processing visualization of large datasets possible in the 1980s, leading to visualization systems like AVS, IBM Data Explorer, and Silicon Graphics IrisExplorer.
+These systems and others forged a visualization paradigm in which non-sophisticated users, i.e., non-programmers, could use graphical user interfaces to develop specific visualizations of their data and provide real-time interactive data exploration.
 
-\section{Methodology}
-\begin{itemize}
-  \item tie into existing VTK-write path
-  \item Galaxy expand to handle unstructured mesh
-  \item tracking particles across domain boundaries
-\end{itemize}
+In the early days, such high-performance visualization engines were expensive and capable only of low-level rendering functions and were limited to dedicated visualization systems.
+This separation of simulation and visualization led to a distinction between general-purpose computation systems for running simulations and graphics-capable systems for visualization.
+Based on a \textit{post-processing} workflow paradigm, simulations would run first on general-purpose systems, then transferred to separate systems for visualization.
+Post-processing remains the standard approach to visualizing large-scale data.
+
+Today, simulation have outstripped the capabilities of even the most heavyweight conventional processors, leading to the development of today's supercomputers - \textit{clusters} of conventional processors (referred to as \textit{nodes}) with very high-speed intercommunications, requiring simulation applications (such as MPM) incorporate sophisticated algorithms to distribute the workload over many cooperating nodes.
+To keep up, visualization systems have also adopted a distributed-parallel architecture; popular general-purpose systems (notably Paraview and VisIt) have been adapted to distributed-memory parallel systems to provide visualization capabilities for simulation systems to run on the largest systems in the world.
+
+\subsection{In-situ Visualization}
+
+Unfortunately, the I/O capabilities of such systems have not matched their computational capabilities, making it even more expensive to export data from the simulation system memory to secondary storage and subsequently read it back into visualization system memory.
+This leads to a tension between the expense of saving time-step data and the requirements of visualization.
+
+\textit{In-situ} visualization is a broad approach to processing simulation data in real-time - that is, \textit{wall-clock} time, as the simulation is running.
+Generally, the approach is to provide \textit{data extracts}, which are condensed representations of the data chosen for the explicit purpose of visualization and computed without writing data to external storage.
+Since these extracts (often images) are vastly smaller than the raw simulation itself, it becomes possible to save them at a far higher temporal frequency than is practical for the raw data, resulting in substantial gains in both efficiency and accuracy.
+In-situ visualization allows simulations to export complete datasets only at the temporal frequency necessary for economic checkpoint/restart.
+
+Many approaches to in-situ visualization exist.
+In \cite{Childs:IJHPCA}, Childs et al.
+provide an exhaustive taxonomy of approaches, differentiating along six axes: the type of integration between simulation and visualization components;  how the components are situated on computing resources, how the visualization components access simulation data, timing, control, and output type.
+For a complete discussion, please refer to the paper.
+The paper also offers a survey of existing in-situ systems.
+Both Paraview and VisIt provide in-situ capabilities; each provides an interface (Catalyst and libsim) with which data is transferred from the simulation code to co-resident visualization components; the simulation code then transfers control to the visualization components, resuming only when the visualization components return.
+
+\subsection{Rendering Techniques}
+At run-time, visualization consists of two phases: converting raw data to renderable primitives (generally triangles) and transferring data to a rendering subsystem that produces pixels in an image buffer.
+Traditionally, this rendering process uses the \textit{rasterization} and Z-buffer algorithms to generate pixels.
+This approach, implemented in hardware, considers each primitive independently, determining which pixels are affected by the primitive, and for each such pixel, it evaluates how far it is from the viewpoint.
+If a given is the closest so far (by comparing depth with the current value in the Z-buffer), it is shaded, saved to the image buffer, and its depth to the Z-buffer.
+Since this algorithm touches all the primitives, the cost is O(n), where n is the number of primitives.
+
+More recently, this approach is being replaced by \textit{ray-tracing}.
+In this algorithm, the rendering loop is across all pixels in the image, asking which primitive is visible at the pixel.
+This algorithm relies on an O(log n) search for each pixel and, on today's hardware (both general-purpose multi-core vector CPUs and graphics cards), is competitive in speed with rasterization and provides additional capabilities.
+First, ray tracing provides a simple approach to realistic lighting effects - notably shadowing - that are hard to implement in rasterization algorithms.
+Both VisIt and Paraview optionally replace their traditional scan-conversion rendering with ray tracing, utilizing Intel's CPU-based OSPRay and Nvidia's GPU-based Optix.
+
+The second advantage of the ray tracing algorithm is that, in some critical cases, it avoids the necessity of reducing data to low-level geometric primitives.
+Critical to the current work is that the rendering of spheres in a rasterization algorithm requires that each sphere be represented as a set of triangles sufficiently large enough to give the appearance of a smooth sphere.
+When there is a large number of spheres to render, rasterization (which recall considers every triangle) becomes very expensive.
+Instead, using ray tracing, we can represent each sphere by its center and radius; rather than exhaustively intersect a ray against each triangle of a sphere's tessellated representation, we can determine the distance between the ray and the center and ask whether it is less than the radius.
+This feature is apparent in Paraview; while the traditional rasterization back end required a filter to create spheres of each point, the points can be rendered directly using the OSPRay back end with substantially improved performance.
+
+\subsection{Parallel Rendering}
+When visualizing on multiple nodes in parallel, the resulting image may contain imagery computed on any participating node.
+To get a correct image, visualization systems generally use one of two algorithms.
+First, they can redistribute the primitives to provide all the information necessary to render a screen patch (possibly the entire image).
+Alternatively, each node renders its portion of the data independently and then uses a distributed-parallel depth-based compositing algorithm to reconcile these partial results into a final image.
+Each has advantages: if the amount of data required to render the image is small, it can be gathered cheaply and then rendered interactively with no further data movement.
+On the other hand, when the data is large, the gather is expensive (and potentially exceeds the memory available on the rendering node(s)), and the post-rendering compositing step is preferable.
+
+When advanced global lighting is required, we note that the second of these algorithms will not work.
+Without shadows, the correct shading for a surface is determined independently of other surfaces and thus can be performed locally before compositing.
+When shadows are required, the surface can be shadowed by a surface \textit{that is not present on current node}.
+Thus post-rendering compositing cannot create a correct image, leaving gathering the data as the alternative.
+
+Galaxy, developed at TACC, differs from OSPRay and Optix in that it is based on a regular partitioning of the computational space, with all the data contained in each partition local to the node responsible for the partition.
+This assumption - derived from the spatial partitioning commonly used in distributed-memory simulation systems - allows Galaxy to do \textit{most} of the work locally, considering (on average) only 1/N of the data, and transferring rays in a nearest-neighbor pattern only when rays (primary and secondary) exit one partition and enter another.
+Rays are retired when they are resolved, and results are transferred to a potentially remote frame buffer.
+
+\subsection{Providing Interactive Visualization In-Situ}
+Of course, in-situ visualization largely precludes exploratory visualization: the in-situ analysis will only produce the anticipated results at set-up time.
+With post-processing, the entire dataset is available when the user sits down in front of the visualization system; the user can adjust viewpoints and parameters interactively during his session, a powerful capability lost in in-situ visualization.
+However, Ahrens et al.
+\cite{Ahrens2014cinema} observed that, for the cost of writing a single timestep of the simulation data, we could save many \textit{images} of the timestep.
+Therefore, we can set up an in-situ process that generates as many images of the data - varying viewpoints and parameters - as the user anticipates and wants.
+Ahrens et al.
+have developed a database format storing these images, and Cinema - a browser for the interactive exploration of such databases.
+We can, for example, specify a range of viewpoints in angle and distance and generate an image for each timestep; the Cinema browser will then present an interactive window giving the appearance of being able to interact with the data in 4 dimensions in real-time.
+
+\section{Study Detail}
+The 3D Oso landslide region of interest spans 1500 x 1500 x 270 m.
+This domain is discretized into 4.2 million material points and a background grid of 8 x 8 x 8 m.
+In most regions, we use two material points in each direction (8 material points per cell), while the sliding area has a higher resolution with four material points in each direction (64 points per cell).
+A Mohr-Coulomb material model with a Total Stress Approach is used to simulate the undrained runout evolution of the Oso landslide in MPM.
+
+In this work, we consider in-situ visualization for MPM.
+In particular, we note that the data for visualization is simple: a large set of points with scalar attributes (currently ~4.2M and potentially billions), scattered among a large number of parallel processes.
+The objective is to generate potentially many images at an appropriate temporal frequency and minimize the need for MPM to dump timestep data.
+We would like to do so with minimal impact on both the MPM codebase (minimizing additional code and dependencies) and the run-time, minimizing the impact on MPM's run-time memory footprint and performance.
+Finally, we would like this interface to be intermittent, allowing us to connect a visualization to a running instance of MPM, enabling us to examine the current state of the simulation.
+
+We chose Galaxy, our ray-tracing-based visualization system, to interface to MPM.
+Galaxy has several advantages that make it attractive in this use; using the compact representation of spheres as the center with radius, rendering a significant number of spheres efficiently, and scaling as necessary to provide the desired performance.
+Ideally, we would like to balance the performance of the MPM simulation with the Galaxy visualization; by developing an M:N point-to-point interface between MPM worker processes and Galaxy processes, we can vary the number of processes in each so that they run at approximately the same pace.
+When out of balance, we can choose to drop frames of the visualization by having MPM continue without pausing if Galaxy is not ready, or not - by having MPM wait for Galaxy to \textit{be} ready.
+
+\subsection{The Interface}
+
+We use a sockets-based interface for this preliminary work, enabling the two components to communicate via the IP protocol.
+While this enables us to position the components anywhere with IP connectivity, by placing the two on the sample supercomputer, we are able to use the full parallel bandwidth of the supercomputer's interconnect to communicate between MPM and Galaxy worker processes.
+
+We use a single connection between the root processes of MPM and Galaxy to control the overall process.
+When Galaxy is launched, its root process creates a master server socket and waits for the simulation's root process to connect.
+Each worker process opens a server socket and waits for zero or more simulation workers to connect.
+
+Periodically, the MPM root process (at its synchronization point) attempts to connect to this server based on a host/port pair given in its configuration file (we generally leave this as localhost/1900 and use an SSH tunnel to connect that local port to wherever the Galaxy root process is running).
+A small amount of control information is passed when the connection is established, including a list of host/port pairs for listening Galaxy worker processes and the number of MPM worker processes transferring data.
+This lets each MPM worker know where to send its data, and each Galaxy worker knows how many connections to expect.
+Data transfer occurs in parallel, and MPM continues when the transfers are complete.
+
+\subsection{Data}
+
+We have implemented an ad-hoc data structure for passing data from MPM to Galaxy for this preliminary work.
+This format consists of a small header containing the name of the variable being transferred and the number of points/data values, followed by a block of binary floats comprising the points, followed by the binary float scalar data values for each point.
+
+\subsection{Galaxy Side}
+
+Once data has arrived on the Galaxy side, it is re-distributed into a Galaxy-compliant partitioning of the computational space.
+As mentioned above, Galaxy relies on a regular subdivision of space - which is not the case for data acquired from MPM, particularly in that multiple MPM workers can send data to each Galaxy node.
+Since each Galaxy worker knows the overall partitioning, this reshuffle is performed using point-to-point messaging via MPI.
+Each worker reads and writes approximately 1/N of the data after examining 1/N of the data.
+We note that while this repartitioning accrue cost, this cost is amortized across all visualization time steps~\cite{abram2018galaxy}.
+
+\subsection{Specifying the Visualization}
+
+Galaxy uses a JSON text file (a \textit{state} file) to specify a visualization.
+The JSON consists of several sections describing: the data objects, the distinct visualizations, listing the data objects and operators to be applied to the lighting environment, a list of cameras, and renderer parameters.
+Names in the data object section correspond to the variables in MPM; at setup time, Galaxy creates empty distributed containers for each named object.
+At run time, the results of reshuffling data received from MPM replace the contents (if any) of these distributed data objects.
+
+From the state file, Galaxy generates a \textit{RenderingSet}, consisting of a list of individual renderings, one for each visualization/camera pair.
+Executing the RenderingSet causes the renderings to be initiated simultaneously, and individual image files are generated.
+
+Galaxy has a state-file generator program for Cinema.
+A \textit{.cinema} file is similar to that specifies \textit{ranges} of parameters, \textit{cinema2state} generates a single state file that contains a list of the visualizations described in the input file with every combination of parameter settings, and a list of cameras with all the combinations of viewpoint parameters.
+Galaxy will generate a complete Cinema database with an interactor for each varying parameter and time.
 
 \section{Results}
-  - show some pretty pictures
-  - provide some runtime stats
+\begin{figure}[!tbp]
+    \includegraphics[width=\linewidth]{figs/timestep-0000.png}
+    \caption{Oso Landslide early}
+    \label{fig:oso-early} 
+\end{figure}
+\vspace{0.2cm}
+\begin{figure}[!tbp]
+  \centering
+    \includegraphics[width=\linewidth]{figs/timestep-0253.png}
+    \caption{Oso Landslide late}
+    \label{fig:oso-late}
+\end{figure}
+
+Given a checkpoint restart file at step 50,000 of the MPM Oso simulation, we ran two versions of the next 15,000 steps to evaluate the cost of in-situ visualization on the simulation.
+Using four nodes of Stampede2 at TACC, each running four MPM worker processes, we first ran 15,000 steps without in-situ visualization but saving checkpoint files every 500 steps.
+In this case, the wall-clock time required to run each 5,000 steps - as determined by timestamps on output files - was approximately 2 hours, 23 minutes.
+
+We then reran the same section of the simulation using the MPM/Galaxy interface to transfer data from MPM to Galaxy every 100 steps and using Galaxy to generate the first two and later three images of each timestep (having noted the runout spread outside our initial camera views) of each step.
+Figures \ref{fig:oso-early} and \ref{fig:oso-late} show two views of this run.
+
+In this case, 
+we ran Galaxy using four worker processes on a single node, so there was a 4:1 ratio of both sending processes to receiving process and sending nodes to receiving nodes.
+Galaxy required approximately 3 seconds for each timestep to receive the data, repartition it, and render the visualizations - far less than the ~three minutes required by MPM to run through 100 steps.
+In this second case, the wall-clock time needed by MPM to run through 5,0000 steps was approximately 2 hours, 26 minutes, or 2 percent more than the first, non-visualized case.
 
 \section{Conclusion}
-- Future work
+
+The above demonstrates that, at this scale, the time cost of in-situ transfer of data from this simulation to a visualization application located on a separate partition of the same supercomputer is negligible.
+Further, the time required to generate the three simple views of the data for every 100 timesteps is much less than the cost of simulating 100 timesteps, indicating that more elaborate visualization (for example, generating many views of each timestep, enabling Cinema-based, real-time interaction the visualization) is achievable without impacting the simulation.
+
+We realized that a new view was necessary to capture the full extent of the run-out at mid-run.   We were able to re-parameterize the visualization on the fly to add this additional view.
+
+\subsection{Future work}
+
+Clearly, the test example we have worked with - the Oso landslide at a relatively coarse resolution does not tax our system.
+However, as observed above, regional-scale simulation of landslides resulting in kilometer-scale runout will require billions of particles rather than millions.
+We believe our approach - notably using a scalable visualization component (Galaxy, \cite{abram2018galaxy}) and a scalable N:M interface architecture enables us to re-scale the visualization process as necessary to achieve in-situ visualization of simulation data with minimal impact on the simulation process.
 
@@
 \section*{Acknowledgment}
-The authors would like to thank...
+The simulations and in-situ visualization of the Oso landslide were conducted on the Texas Advanced Computing Center’s (TACC’s) cluster Stampede2.
+Krishna Kumar would like to thank the U.S.
+National Science Foundation grant CMMI-2022469.
+Greg Abram would like to thank the U.S.
+Department of Energy grant DE-SC0012513.
+However, any opinions,
+findings and conclusions or recommendations expressed in this material are those of the authors and do not necessarily reflect the views of the National Science Foundation.
 

### 2022-02-08 response-reviewers.tex
Revised manuscript

% LaTeX rebuttal letter example. 
% Licensed under cc by-sa 3.0 with attribution required.

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{lipsum} % to generate some filler text
\usepackage{fullpage}

% package needed for optional arguments
\usepackage{xifthen}
% define counters for reviewers and their points
\newcounter{reviewer}
\setcounter{reviewer}{0}
\newcounter{point}[reviewer]
\setcounter{point}{0}

% This refines the format of how the reviewer/point reference will appear.
\renewcommand{\thepoint}{P\,\thereviewer.\arabic{point}}

% command declarations for reviewer points and our responses
\newcommand{\reviewersection}{\stepcounter{reviewer} \bigskip \hrule
                  \section*{Reviewer \thereviewer}}

\newenvironment{point}
   {\refstepcounter{point} \bigskip \noindent {\textbf{Reviewer~Point~\thepoint} } ---\ }
   {\par }

\newcommand{\shortpoint}[1]{\refstepcounter{point}  \bigskip \noindent 
	{\textbf{Reviewer~Point~\thepoint} } ---~#1\par }

\newenvironment{reply}
   {\medskip \noindent \begin{sf}\textbf{Reply}:\  }
   {\medskip \end{sf}}

\newcommand{\shortreply}[2][]{\medskip \noindent \begin{sf}\textbf{Reply}:\  #2
	\ifthenelse{\equal{#1}{}}{}{ \hfill \footnotesize (#1)}%
	\medskip \end{sf}}

% References
\usepackage{biblatex}
\addbibresource{references.bib}

% Title content
\title{Revisions for ``In-situ visualization of natural hazards with Galaxy and Material Point Method''}
\date{}
\begin{document}
\maketitle
We thank the reviewers for their critical assessment of our work. We have substantially revised our paper improving the readability of the methodology sections and accessible visualizations as well. 
In the following, we address their concerns point by point.

\section*{Editor-in-Chief comments}

\begin{point}
I agree with the reviewer who says that your claim of “can achieve in-situ viz … with billions of particles” needs to be justified, given that the demo in the paper only uses 4 million.
\end{point}
\shortreply{\label{Expl:scaling}
Thank you for the comment. We have added strong scaling results for the MPM code as well as added some published results that show the capability of Galaxy to simulate a 1 billion particle model of Cosmic Web\cite{abram2018galaxy}. Having said that we have also toned down our claims for doing billion particle simulations.}

\begin{point}
Also agree that M:N point-to-point is unclear for readers. On page 4, the paragraph starting “We choose Galaxy…” needs expansion and more careful explanation of Galaxy for the non-insider. The whole paragraph currently is quite impenetrable.
\end{point}
\begin{reply}
\label{Expl:MN}
Thank you for the comments. We have completely rewritten the paragraph explaining in detail how the M:N communication works.

\begin{quote}
Two concerns arose in developing the MPM-Galaxy interface.   First, we would like to balance the performance of the MPM simulation with the Galaxy visualization by varying the ratio of MPM processes to Galaxy processes.   Second, the spatial partitioning layout in MPM is not absolute; while each MPM process manages a relatively compact region of space, these regions overlap and do not form the clean Cartesian boundaries necessary for Galaxy.   To address these problems, we developed an asynchronous M to N data transfer scheme between M MPM processes and N Galaxy processes that incorporates a sorting step.   This is implemented by communicating an appropriate layout of partitions, given the number of Galaxy processes in play and the bounding box of the MPM computational space, to MPM at data transfer time.   Each MPM process then traverses its particle data, determines which Galaxy process's partition space the particle lies, and transfers them accordingly.  The M:N point-to-point interface between MPM worker processes and Galaxy processes offers better load balancing performance.  We can vary the number of processes for MPM and Galaxy so that they run at approximately the same pace.  When out of balance, we can
choose to drop frames of the visualization by having MPM
continue without pausing for Galaxy (if Galaxy is not ready), or by having MPM wait for Galaxy to be ready.
\end{quote}
\end{reply}

\begin{point}
The Result section does need some elaboration. You are comparing one run with checkpointing every 5,000 steps but no in-situ visualization, to one that does visualization in-situ every 100 steps. In the first case, data exploration would only be available from post-processing the saved data at checkpoint times. In the second, data visualizations are available with much greater frequency. For scientific findings, would data at checkpoint times be sufficient? Or would more frequent I/O be needed? It sounds like you need to expand in your discussion section on the difference in interpretation potential of these two cases, from the science point of view. You also need to expand the final section where you extrapolate the need to the scale of billions of particles. How do you arrive at that estimate? What makes you think that your system will be able to achieve that scale? Please spend a little more effort persuading readers here, given that the paper is short and you have the space.
\end{point}
\shortreply{Regional-scale simulations are extremely time consuming both in the sense of running the simulations as well as the post-hoc visualization of the results. When the simulation scale increase, writing all the results at a fixed frequency further increases the computational time. Due to the I/O bottleneck, often researchers minimize the amount of data written, while also increasing the duration between data outputs.

We agree that we haven't scaling results. We have added the MPM scaling results to 128 nodes and included references for the claim of Galaxy being able to simulate billions of particles running on 128 nodes~\cite{abram2018galaxy}. We have also revised our claims accordingly.

We have also added the following statement to qualify why we need in-situ viz for regional-scale landslide: 
\begin{quote}
    In-situ visualization with Galaxy/MPM integration offers insights into the dynamics of very fast runouts, such as the Oso landslide (total duration of runout is 120s) with a very fine temporal resolution, which are otherwise infeasible using traditional post-processing visualization.
\end{quote}
}

\shortpoint{p.1, col.2, l.27: focuses $->$ focus \dots provides $->$ provide}
\shortreply{Thanks. We have fixed this issue.}

\shortpoint{line 37: refer to [5] - do not use references as parts of speech; note that the article layout will set these as superscripts; rephrase with something like ``refer to Kumar et al. [5]"}
\shortreply{Thank you we have revised all references in the text accordingly.}

\shortpoint{line 43: Scaling \dots require $->$ requires}
\shortreply{Thanks. We have fixed this issue.}

\shortpoint{p,2, col.2, l.54: simulation have $->$ simulations have}
\shortreply{Thanks. We have fixed this issue.}

\begin{point}
– You will need to add short author bios at the end. Here is the way that these should be set up: 
//Author full name// is a //role// in the //department, institution, country, post code//. //His/Her/Their// research interests include //list 3 to 5 brief topics//. //Last name// received a //highest degree// in //field/subject/area// from //institution//. [Do not list year.] //He/She/They// are a member //or higher// of //relevant professional associations [we like to list IEEE or IEEE CS first, but it's fine to group by type of membership to keep it brief]//. //If this person is a CISE EB member or a Computer Society elected officer, you can mention this separately here.// Contact //him/her/them// at [best email address]. 
\end{point}
\shortreply{Thanks. We have added the bios for all the authors.}

\section*{Associate Editor}
\shortpoint {My main concern with the current state of the manuscript is that scalability is not discusses in detail (and certainly not empirically demonstrated) and the mapping of processes (both simulation and visualization processes) onto physical computing resources is far from clear. Better definition for "develop a scalable N:M interface architecture"}
\shortreply{Thank you for the comments. We have added scaling results and references to show the scalability of both MPM and Galaxy. Also, we have revised the M:N interface description. Please see our responses to the Editor P0.1 and P0.2.}

\shortpoint{ better support for "The developed approach can achieve in-situ visualization of regional scale landslides with billions of particles with minimal impact on the simulation process."}
\shortreply{\label{Expl:scaling}
Thank you for the comment. We have added strong scaling results for the MPM code as well as added some published results that show the capability of Galaxy to simulate a 1 billion particle model of Cosmic Web\cite{abram2018galaxy}. We have also quantitatively show the amortized simulation time impact increased by only 2\% compared to no-visualization which is an extreme case. Typically, post-processing visualization is done every 1000 steps, which significantly slows down the runtime by more than 2\% observed in in-situ visualization.}

\shortpoint{better description of Galaxy architecture and use of OSPRay / Embree hardware accel}

\begin{reply}
Thank you for the comments. We have added a description of how Galaxy links to OSPRay and Embree and added a reference of ~\cite{abram2018galaxy} which provides detailed description of Galaxy implementation. 
\begin{quote}
    Note that Galaxy is actually a framework for the management and exchange of rays between processes; it does not actually perform ray/surface or ray/volume     intersections itself.  Rather, it relies on OSPRay (or, rather, OSPRay's Embree and OpenVKL underpinnings) to perform the \textit{local} ray tracing.  This could equally well be performed by Optix.  For more information on Galaxy, please see Abram et. al. \cite{abram2018galaxy}.    
\end{quote}
\end{reply}

\shortpoint{better description of overhead / resource costs }
\begin{reply}
Thank you we have rewritten the results section on the costs and overheads.

\begin{quote}
    Using four nodes of Stampede2 at TACC, each running four MPM worker processes, we first ran 15,000 steps without in-situ visualization but saving checkpoint files every 5,000 steps, resulting in five time step datasets over this interval. In this case, the wall-clock time required to run each 5,000 steps - as determined by timestamps on output files - was approximately 2 hours, 23 minutes.

    We then reran the same interval of the simulation with the same checkpoint dumps, but now using the MPM/Galaxy interface to transfer data from MPM to Galaxy every 100 steps (0.1 s of real-time runout, where a difference in the runout can be perceived).  In-situ visualization with Galaxy/MPM integration offers insights into the dynamics of very fast runouts, such as the Oso landslide (total duration of runout is 120s) with a very fine temporal resolution, which are otherwise infeasible using traditional post-processing visualization. We found that Galaxy was substantially faster than MPM, even when running on a single node.  To make full use of the I/O bandwith available on that node, we use a M:N interface of 4:1 with four MPM worker nodes to one Galaxy worker nodes.  With this setup, Galaxy required approximately 6.5 seconds for each timestep to receive the data, repartition it, and render the visualizations - far less than the three minutes required by MPM to run through 100 steps.   In this second case with in-situ visualization, the wall-clock time needed by MPM to run through 5,000 steps was approximately 2 hours, 26 minutes, or 2 percent more than the first, non-visualized case. 
\end{quote}

\end{reply}

% Let's start point-by-point with Reviewer 1
\reviewersection

\begin{point}
My main concern with the current state of the manuscript is that scalability is not discusses in detail (and certainly not empirically demonstrated) and the mapping of processes (both simulation and visualization processes) onto physical computing resources is far from clear.
\end{point}
\shortreply{\label{Expl:scaling}
Thank you for the comment. We have added strong scaling results for the MPM code as well as added some published results that show the capability of Galaxy to simulate a 1 billion particle model of Cosmic Web\cite{abram2018galaxy}.}

\begin{point}
the abstract states "In this study, we develop a scalable N:M interface architecture to visualize regional-scale landslides.": what is N and what is M? Even though some clues appear latter in the text, the reader has not , yet, any idea of what is being said.
\end{point}
\shortreply{Thank you for the comment. We have completely revised this section. Please see our response to the Editor P0.1 and P0.2.}

\begin{point}
 lines 5..7 of page 4 state: "allows Galaxy to do most of the work locally, considering (on average) only 1/N of the data,"; N hasn't been defined yet. I suppose we can infer that N are the number of partitions of the computational space, has decided by Galaxy, but this needs to be clarified. Even because later N seems to be the number of Galaxy processes, and it is far from obvious that there should be a 1:1 correspondence between space partitions and processes...
\end{point}
\begin{reply}
\label{Expl:MN}
Thank you for the comments. We have completely rewritten the paragraph explaining in detail how the M:N communication works. Please see our response to the Editor P0.1. We have also moved the discussion in section D on parallel rendering to later in in the Galaxy Side section after M:N interfaces are introduced. 
\end{reply}

\begin{point}
 lines 7..10 of the 2nd column of page 4 state that "by developing an M:N point-to-point interface between MPM worker processes and Galaxy processes, we can vary the number of processes in each so that they run at approximately the same pace.". This is the 1st time the reader is given an hint that M are (I suppose) the number of simulation processes and N the number of visualization (Galaxy) processes (and simultaneously the number of Galaxy partitions of the space). Also it is vaguely suggested that by varying the relationship between M and N some kind of synchronicity (or maybe load balancing) between the simulation and the visualization can be achieved. But this must be clarified.
\end{point} 
\shortreply{Apologies for the vague description. We have completely rewritten this section. Please see our response to Editor P0.1.}

\begin{point}
line 23 of the 2nd column of page 5 states "In this case, we ran Galaxy using four worker processes on a single node, ".  Now the issue of mapping processes to nodes becomes suddenly obvious and it has not been explained. What are M and N after all? The number of processes for sim and viz or the number of nodes ? It is also not clear whether additional nodes are used for visualization (compared to the non in-situ case). is this an extra node, not used by the MPM processes? It certainly seems so, but it should be made clearer.
\end{point}
\begin{reply}
Apologies for the lack of clarity. M:N interface refers to M processors running MPM and N processors running Galaxy. For comparing the case of no visualization (4 Nodes for MPM) with Galaxy in-situ visualization (1 additional node for Galaxy, in addition to 4 MPM tasks). This explanation has been added to the text.

Offering in-situ visualization capabilities incurs additional compute cost at run-time. However, post-processing visualization also has this addition compute cost, which is likely to be higher than in-situ visualization due to increased storage and latency requirements. We have added the above statements to our manuscript. 
\end{reply}

\begin{point}
Also I wonder how would this scale. Is there any notion of how should N grow with M for larger simulations and how would this impact on execution time? In other words, how do both systems (MPM and Galaxy) scale with respect to each other and the problem size?
\end{point}
\shortreply{Thank you for your comments. We agree that we have not demonstrated the integrated approach to scale to billions of particles. However, we have included strong-scaling results of MPM running of 128 nodes. In addition, we have added citations to back-up our claim of Galaxy supporting a billion particle simulations of Cosmic Web~\cite{abram2018galaxy}. Since the scaling of both Galaxy and MPM are both linear to 128 nodes, we expect the M:N ratio would be similar to 4:1 with the 4 million material points. However, as simulations become more computationally intensive, we would have to run some trials to understand how this ratio will change. We agree with the reviewers, we have toned down our claim to scale to billions of particles.}

\begin{point}
This scalability issue is important because the last sentence of the abstract states : "The developed approach can achieve in-situ visualization of regional scale landslides with billions of particles with minimal impact on the simulation process." Is there any data supporting this claim? The presented study seems to use 4.2 million points (line 36, column 1, page 4). It is not clear how many processing nodes have been used (clearly M=4, but how many nodes assuming there are sim processes, and a single node is used to run 4 viz processes). So how can one conclude on the scalability to billions of points or even on the scalability to larger computing systems? I read reference [8] and can see results on Galaxy scalability up to 128 nodes; but I couldn't establish a clear relationship with this M:N ratio. Nevertheless, given the nature of this journal, I think the manuscript should be as self contained as possible. In this sense, I would like the authors to make this scalability issue more clear and/or rephrase the claim on the last sentence of the abstract and last on section V.A ("Future work").
\end{point}
\shortreply{Thank you for comments. We agree with the reviewer. As explained in the previous reply, we have added some scaling results. We also acknowledge the claim is not demonstrated and we have changed the claim accordingly.}

\reviewersection
\begin{point}

For a submission on Hardware-accelerated Ray Tracing for Scientific Simulations, I expected hardware accelerated ray tracing to play a larger part in the paper. I believe the content is highly relevant to the issue, but the paper needs some minor messaging adjustments. For example, "Galaxy, developed at TACC, differs from OSPRay and Optix in that it is based on a regular partitioning of the computational" I found this paragraph awkward and hard to parse the difference. OSPRay and Optix work on local data and Galaxy builds on top of hardware accelerated ray tracing (i.e., OSPRay) to work a distributed memory setting. Asynchronous distributed memory is hard, and worthy of discussing. I had to go to the github repository to discover that Galaxy leverages hardware accelerated ray tracing through OSPRay.
\end{point}
\shortreply{Thank you for the comment. We have added the following text:
\begin{quote}
    Note that Galaxy is actually a framework for the management and exchange of rays between processes; it does not actually perform ray/surface or ray/volume intersections itself.   Rather, it relies on OSPRay (or, rather, OSPRay's Embree and OpenVKL underpinnings) to perform the \textit{local} ray tracing.  This could equally well be performed by Optix.   For more information on Galaxy, please see Abram et. al. \cite{abram2018galaxy}.
\end{quote}
}

\begin{point}
One note on the cost of the system. While the additional runtime was only 2\% more than running without Galaxy, isn't there an additional resource cost? For example, if we calculate the costs in terms of node hours consumed by the job then with an additional node used for visualization the cost is greater than 2\%. Additionally, the abstract states that the system can achieve visualizations of billions of particles with minimal impact on the simulation, but it was not demonstrated. What was demonstrated was 3 orders of magnitude less. While theoretically this might be possible, the statement should be backed-up. This language is toned down in the conclusion to "we believe".
\end{point}
\shortreply{Thank you for the comment. We acknowledge the lack of scaling results. We have added the MPM strong scaling to 128 nodes and Galaxy also shows good scaling to 128 nodes as demonstrated by Abram~\cite{abram2018galaxy}. Yes, we acknowledge that in-situ visualization with Galaxy incurs additional compute nodes. By tuning the M:N ratio we achieve the runtime for both Galaxy and MPM to be the same. Although there is an increase in computational resource, we aim for both simulation and visualization engines to not wait on each other. We agree that the amortized cost of 2\% does not include the additional cost of nodes for visualization. However, the runtime comparison is to show the choice of using in-situ visualization does not significantly increase the cost of visualizing the results.

We agree that offering in-situ visualization capabilities incurs additional compute cost at run-time. However, post-processing visualization also has this addition compute cost, which is likely to be higher than in-situ visualization due to increased storage and latency requirements. We have added the current paragraph to the manuscript.}

\begin{point}
"We realized that a new view was necessary to capture the full extent of the run-out at mid-run. We were able to re-parameterize the visualization on the fly to add this additional view." Is this a conclusion of the paper? 
\end{point}
\shortreply{Apologies. We have added the explanations of adding views on the fly. We have removed the statement from the conclusion.}

\begin{point}
- large set of points with scalar attributes (currently 4.2M and potentially billions)

Change to:

large set of points (currently 4.2M and potentially billions) with scalar attribute.
\end{point}
\shortreply{Thank you. We have made the suggested change.}

\shortpoint{This paper doesn't fit the style of the examples that were provided by the editors. Its formatted as a technical paper, although I found that the content wasn't overly technical in most places. Details such as "we generally leave this as localhost/1900 and use an SSH tunnel to connect that local port to wherever the Galaxy root process is running" is probably too much that doesn't add much to the understanding of the contribution.}
\shortreply{We agree with the reviewer and the statement has been removed.}

\shortpoint{Improve the introduction.}
\shortreply{Thank you for the comment. We have revised the introduction to make it clear.}

\reviewersection

\begin{point}
The topic of this manuscript is of a great importance to all readers who are interested in exascale simulations and visualization in the STEM field. The I/O operations during the simulation is always a burden for HPC applications, to avoid such problem, in-situ visualization is needed and should be without extra time cost. The authors coupled the MPM with Galaxy to carry out the in-situ visualization for the Oso landslide with a negligible time cost. The manuscript discussed how exascale simulations needs rendering engines capable of rendering billions of light paths. Most rendering engines only sample a limited number of random light paths at each time step producing images with low visual quality. Therefore, this study used TACC Galaxy Ray tracing engine which offers distributed asynchronous in-situ visualization capabilities for exascale simulations. 
\end{point}
\shortreply{Thank you for your encouraging comments.}
\printbibliography
\end{document}

### 2022-02-08 bare_jrnl.tex
Revised manuscript

@@
 
+\usepackage{subcaption}
 \usepackage[pdftex]{graphicx}
@@
 \usepackage{url}
+\usepackage{graphicx, float}
+\usepackage{caption,lipsum}
 
@@
 
-\title{In-situ visualization of natural hazards with Galaxy and Material Point Method}
-\author{Greg Abram, %~\IEEEmembership{Fellow,~OSA,}
-        Andrew Solis,
-        Yong Liang,
-        and~Krishna Kumar% <-this % stops a space
-\thanks{Department of Civil, Architectural and Environmental Engineering, University of Texas at Austin, TX, 78712 USA e-mail: (krishnak@utexas.edu).}% <-this % stops a space
-\thanks{Texas Advanced Computing Center, University of Texas at Austin, TX,.}% <-this % stops a space
-\thanks{University of California at Berkeley, CA,.}% <-this % stops a space
-\thanks{Manuscript received September 3rd, 2021.}}
+\title{In-situ visualization of regional-scale natural hazards with Galaxy and Material Point Method}
+
+\author{Greg Abram$^1$, %~\IEEEmembership{Fellow,~OSA,}
+        Andrew Solis$^1$,
+        Yong Liang$^2$,
+        and~Krishna Kumar$^{3}$% <-this % stops a space
+\thanks{$^1$Texas Advanced Computing Center, University of Texas at Austin, TX,.}% <-this % stops a space
+\thanks{$^2$University of California at Berkeley, CA,.}% <-this % stops a space
+\thanks{$^{3}$Department of Civil, Architectural and Environmental Engineering, University of Texas at Austin, TX, 78712 USA e-mail: (krishnak@utexas.edu).}% <-this % stops a space
+\thanks{Manuscript received September 3rd, 2021. Revised on January 31st, 2022.}}
 
@@
 
-% make the title area
+%\maketitle
+
 \maketitle
@@
 \begin{abstract}
-Visualizing regional-scale landslides is the key to conveying the threat of natural hazards to stakeholders and policymakers.
-Traditional visualization techniques are restricted to post-processing a limited subset of simulation data and are not scalable to rendering exascale models with billions of particles.
+Visualizing regional-scale landslides is essential to conveying the threat of natural hazards to stakeholders and policymakers.
+Traditional visualization techniques are restricted to post-processing a limited subset of simulation data and are not scalable to rendering regional-scale models.
 In-situ visualization is a technique of rendering simulation data in real-time, i.e., rendering visuals in tandem while the simulation is running.
-In this study, we develop a scalable N:M interface architecture to visualize regional-scale landslides.
+This study develops a scalable N:M interface architecture to visualize regional-scale landslides.
 We demonstrate the scalability of the architecture by simulating the long runout of the 2014 Oso landslide using the Material Point Method coupled with the Galaxy ray tracing engine rendering 4.2 million material points as spheres.
 In-situ visualization has an amortized runtime increase of 2\% compared to non-visualized simulations.
-The developed approach can achieve in-situ visualization of regional-scale landslides with billions of particles with minimal impact on the simulation process.
+The developed approach can achieve in-situ visualization of regional-scale landslides with minimal impact on the simulation process.
 \end{abstract}
+
+%\begin{figure*}
+%\includegraphics[width=\linewidth]{figs/teaser.png}
+%\caption{Galaxy renderings of the Oso landslide early(L) and late(R) in the runout. The color gradient shows the amount of particle displacements from %their original position.}
+%\label{Fig:runout}
+%\end{figure*}
+
+ \begin{figure*}[!tbp]
+    \begin{subfigure}[b]{0.333\textwidth}
+        \includegraphics[width=\textwidth]{figs/tstep-010000_00000.png}
+        \caption{runout after 10 s}
+        %\label{fig:oso-10} 
+    \end{subfigure}
+    \begin{subfigure}[b]{0.334\textwidth}
+        \includegraphics[width=\textwidth]{figs/tstep-015000_00000.png}
+        \caption{runout after 15 s}
+        %\label{fig:oso-15} 
+    \end{subfigure}
+    \begin{subfigure}[b]{0.333\textwidth}
+        \includegraphics[width=\textwidth]{figs/tstep-020000_00000.png}
+        \caption{runout after 20 s}
+    \end{subfigure}
+    \caption{Early stages of the simulated landslide runout visualized using in-situ rendering of CB-Geo MPM simulation with TACC Galaxy.
+The color gradient shows the amount of particle displacements from their original position.}
+    \label{fig:oso3}
+\end{figure*}
 
@@
 The human capacity to deny danger endangers lives and property~\cite{kim2020public}.
-In 2020, twenty-two extreme events cost the United States \$95 billion in damages.
+In 2021, twenty extreme events cost the United States \$145 billion in damages.
 The underlying problem is the uncertainty of the actual event and a lack of understanding of its potential impacts.
@@
 However, these visualizations must be physically sound to be perceived as realistic and accepted as valid.
+\Cref{fig:oso3} shows a realistic rendering of an early-stage landslide runout.
 Visualization is also context and user-specific, i.e., different users may be interested in different aspects of the experimental or simulation data sets.
@@
 Despite the landslide risks, most regional-scale landslide hazard analyses do not consider downstream impacts and the aerial extent of debris-flow runouts~\cite{USGS2017debris}.
-Traditional numerical methods such as the Finite Element Method (FEM) primarily focuses on the onset of failure but provides limited information on the post-failure runout mechanisms due to mesh distortions at large  deformations~\cite{soga2016trends}.
-Modeling the impacts of the potential runout of large landslides is possible with new tools such as Material Point Method (MPM)~\cite{kumar2019scalable}.
+Traditional numerical methods such as the Finite Element Method (FEM) primarily focus on the onset of failure but provide limited information on the post-failure runout mechanisms due to mesh distortions at large  deformations~\cite{soga2016trends}.
+Modeling the impacts of the potential runout of large landslides is possible with tools such as Material Point Method (MPM)~\cite{kumar2019scalable}.
 MPM is a mesh-free method that discretizes the domain as a collection of \textit{material points} moving on a background grid, and Newton's laws of motion determine their deformations.
-In this study, we employ the CB-Geo MPM code (\url{https://github.com/cb-geo/mpm}), and a typical MPM computation cycle is shown in~\cref{fig:mpm}.
-For more information on MPM and the code implementation, refer to~\cite{kumar2019scalable}.
-MPM results of the landslide runout can be translated into compelling visualizations to inform decision-makers of potential hazards.
+~\Cref{fig:mpm} shows a typical MPM computation cycle.
+In this study, we employ the CB-Geo MPM code (\url{https://github.com/cb-geo/mpm}).
+For more information on MPM and the code implementation, refer to Kumar et.
+al.
+~\cite{kumar2019scalable}.
+MPM simulations of the landslide runout can be translated into compelling visualizations to inform decision-makers of potential hazards (see~\cref{fig:oso3}).
 
-\begin{figure}[!htbp]
-    \centering
+\begin{figure}[tbp]
     \includegraphics[width=\linewidth]{figs/mpm.pdf}
@@
 
-A regional-scale landslide simulation of a kilometer-scale runout with MPM requires billions of material points and millions of cells in the background grid.
-Scaling applications to simulate a billion particles require exascale high-performance computing (HPC) architectures with efficient Message-Passing Interface (MPI).
-This study uses the CB-Geo MPM code, which exploits a hybrid MPI+X approach to achieve regional-scale simulation of landslides.
+An MPM simulation of regional-scale landslide requires millions to billions of material points and background grid nodes.
+Scaling applications to model region-scale hazards requires  high-performance computing (HPC) architectures with efficient Message-Passing Interface (MPI).
+The CB-Geo MPM code exploits a hybrid MPI+X approach to achieve regional-scale simulation of landslides.
 In addition, as the landslide runout progresses, the workload distributed across the compute nodes becomes imbalanced.
 We adopt a distributed graph-partitioning technique to load balance both the background mesh and particle distribution dynamically.
+\Cref{fig:scaling} shows the hybrid MPI+X approach of CB-Geo MPM offers linear scaling up to 10k cores for simulating 7 million material points on Texas Advanced Computing Center (TACC) Stampede2 with Intel Xeon Skylake nodes.
 
-Data retrieval and visualization in High-Performance Computing applications have long been a bottleneck.
-Current techniques for visualizing exascale simulations involve writing a temporal slice of a subset of the data to disk, often at a much coarser resolution than the original data, which leads to a significant portion of information being disregarded and potentially lost, followed by post-processing.
-In practice, this usually means that the data are stored only at specific time steps.
-An exascale simulation would output terabytes of data, resulting in a significant portion of the compute time spent on I/O~\cite{byna2020exahdf5}.
-Nevertheless, the scale of such discrete geometry visualization will push commercial rendering engines like Mantra and Arnold to the limit.
-The amount of data would be several orders of magnitude higher at exascale levels, making the simulation spend most of the compute time in I/O.
+\begin{figure}[!htbp]
+    \centering
+    \includegraphics[width=\linewidth]{figs/scaling.eps}
+    \caption{Strong scaling results of CB-Geo MPM simulating 7 million material points on TACC Stampede2.} 
+    \label{fig:scaling}
+\end{figure}
+
+Data retrieval and visualization in HPC applications have long been a bottleneck.
+Current techniques for visualizing regional-scale simulations involve writing a temporal slice of a subset of the data to disk, often at a much coarser resolution than the original data, which leads to a significant portion of information being disregarded and potentially lost.
+In practice, the simulation data are only stored at specific time steps, often every 1000 or 10,000 steps, followed by post-processing of data.
+A  regional-scale simulation would output up to terabytes of data, resulting in a significant portion of the compute time spent on I/O~\cite{byna2020exahdf5}.
+Nevertheless, such regional-scale discrete geometry visualization will not be efficient or scalable without leveraging distributed computing.
 
@@
 In-situ visualization libraries such as ParaView Catalyst and SENSEI offer the ability to couple existing data models (such as VTK) to query regions of interest and to render simulations.
-However, scaling rendering engines to exascale simulations remains a challenge requiring billions of lightpaths to be rendered.
+However, scaling rendering engines to regional-scale simulations remains a challenge requiring billions of lightpaths to be rendered.
 Most rendering engines only sample a limited number of random light paths at each step.
 Thus, the visual quality of rendered images suffers from estimator variance, which appears as visually distracting noise~\cite{yang2021foveated}.
-Ray tracing engines such as TACC Galaxy~\cite{abram2018galaxy} (\url{https://github.com/cb-geo/mpm}) offer distributed asynchronous in-situ visualization capabilities for exascale simulations.
+Ray tracing engines such as TACC Galaxy~\cite{abram2018galaxy} (\url{https://github.com/tacc/galaxy}) offer distributed asynchronous in-situ visualization capabilities for regional-scale simulations.
 
-This study develops an in-situ visualization technique to simulate and render regional-scale landslides by coupling the CB-Geo MPM code for large-deformation modeling with distributed asynchronous ray tracing engine TACC Galaxy.
-In this study, we apply the in-situ visualization technique in MPM to simulate and visualize the disastrous 2014 Oso landslide, also known as the SR 530 landslide, which occurred in the northwest Washington state in the US, resulting in an 1100 m runout.
+In this study, we develop an in-situ visualization technique to simulate and render regional-scale landslides by coupling the CB-Geo MPM code for large-deformation modeling with distributed asynchronous ray tracing engine TACC Galaxy.
+We apply the developed in-situ visualization technique in MPM to simulate and visualize the disastrous 2014 Oso landslide, also known as the SR 530 landslide, which occurred in the northwest Washington state in the US,  resulting in an 1100 m runout~\cite{yerro2019runout}.
 
@@
     \centering
-    \includegraphics[width=\linewidth]{figs/osofull.png}
-    \caption{Visualization of a timestep of the MPM simulation of the Oso landslide.
-Particles are represented as spheres with a radius approximating the original resolution.
-Color represents the distance of each particle from its original location.}
-    \label{fig:osofull}
+    \includegraphics[width=\linewidth]{figs/sideBySide-Global-Illum.png}
+    \caption{Detail of visualization without (L) and with (R) ambient occlusion and shadow global illumination.} 
+    \label{fig:global_illum}
 \end{figure}
 
-\Cref{fig:osofull} shows a visualization of the final runout of the MPM simulation of the Oso landslide.
-In this case, we use Galaxy to render the scene using spheres to represent the material points in the simulation, each with a fixed radius matching the original particle grid resolution and colored by the distance between each particle's current and original location.
-Galaxy's global lighting model is used to calculate shadows for additional depth perception.
+\Cref{fig:oso3,fig:timing} show visualizations of the MPM simulation at different stages of the Oso landslide runout.
+In this case, we use Galaxy to render the scene using spheres to represent the material points in the simulation, each with a fixed radius matching the original particle grid resolution and colored by the distance between each particle's current and original location (i.e., particle displacement).
+In this study, we use Galaxy's global illumination model - shadows and ambient occlusion - to provide additional depth perception.
+\Cref{fig:global_illum} shows a side-by-side comparison of detail of a timestep visualizations with and without global illumination.
+For a discussion of the impact of these techniques on the ability to discern complex shapes, please see Grottel et.
+al \cite{Grottel12}.
 
@@
 Systems for visualizing large-scale scientific data have been available for decades.
-High-performance graphics engines - add-ons to powerful workstations first made interactive post-processing visualization of large datasets possible in the 1980s, leading to visualization systems like AVS, IBM Data Explorer, and Silicon Graphics IrisExplorer.
+High-performance graphics engines, add-ons to powerful workstations, first made interactive post-processing visualization of large datasets possible in the 1980s, leading to visualization systems like AVS, IBM Data Explorer, and Silicon Graphics IrisExplorer.
 These systems and others forged a visualization paradigm in which non-sophisticated users, i.e., non-programmers, could use graphical user interfaces to develop specific visualizations of their data and provide real-time interactive data exploration.
@@
 
-Today, simulation have outstripped the capabilities of even the most heavyweight conventional processors, leading to the development of today's supercomputers - \textit{clusters} of conventional processors (referred to as \textit{nodes}) with very high-speed intercommunications, requiring simulation applications (such as MPM) incorporate sophisticated algorithms to distribute the workload over many cooperating nodes.
-To keep up, visualization systems have also adopted a distributed-parallel architecture; popular general-purpose systems (notably Paraview and VisIt) have been adapted to distributed-memory parallel systems to provide visualization capabilities for simulation systems to run on the largest systems in the world.
+Today, simulations have outstripped the capabilities of even the most heavyweight conventional processors, leading to the development of today's supercomputers - \textit{clusters} of conventional processors (referred to as \textit{nodes}) with very high-speed intercommunications.
+Super computers require simulation applications, such as MPM, incorporate sophisticated algorithms to distribute the workload over many cooperating nodes.
+To keep up, visualization systems have also adopted a distributed-parallel architecture; popular general-purpose systems (notably Paraview and VisIt) have been adapted to distributed-memory parallel systems to provide visualization capabilities for simulations running on the largest super computing systems in the world.
 
@@
 Unfortunately, the I/O capabilities of such systems have not matched their computational capabilities, making it even more expensive to export data from the simulation system memory to secondary storage and subsequently read it back into visualization system memory.
-This leads to a tension between the expense of saving time-step data and the requirements of visualization.
+Expensive I/O leads to a tension between the expense of saving time-step data and visualization requirements.
 
@@
 Many approaches to in-situ visualization exist.
-In \cite{Childs:IJHPCA}, Childs et al.
-provide an exhaustive taxonomy of approaches, differentiating along six axes: the type of integration between simulation and visualization components;  how the components are situated on computing resources, how the visualization components access simulation data, timing, control, and output type.
-For a complete discussion, please refer to the paper.
+Childs et al.\cite{Childs:IJHPCA} provide an exhaustive taxonomy of approaches, differentiating along six axes: the type of integration between simulation and visualization components,  how the components are situated on computing resources, how the visualization components access simulation data, timing, control, and output type.
+For a complete discussion, please refer to the paper~\cite{Childs:IJHPCA}.
 The paper also offers a survey of existing in-situ systems.
@@
 Traditionally, this rendering process uses the \textit{rasterization} and Z-buffer algorithms to generate pixels.
-This approach, implemented in hardware, considers each primitive independently, determining which pixels are affected by the primitive, and for each such pixel, it evaluates how far it is from the viewpoint.
-If a given is the closest so far (by comparing depth with the current value in the Z-buffer), it is shaded, saved to the image buffer, and its depth to the Z-buffer.
+This approach, implemented in hardware, considers each primitive independently, and iterates over all the pixels to determine which pixels are affected by the primitive and evaluates how far each affected pixel is from the viewpoint.
+If a given affected pixel is the closest so far (by comparing depth with the current value in the Z-buffer), it is shaded and saved to the image buffer, and its depth is saved to the Z-buffer.
 Since this algorithm touches all the primitives, the cost is O(n), where n is the number of primitives.
@@
 This algorithm relies on an O(log n) search for each pixel and, on today's hardware (both general-purpose multi-core vector CPUs and graphics cards), is competitive in speed with rasterization and provides additional capabilities.
-First, ray tracing provides a simple approach to realistic lighting effects - notably shadowing - that are hard to implement in rasterization algorithms.
+Ray tracing provides a simple approach to realistic lighting effects - notably shadowing and ambient occlusion - that are hard to implement in rasterization algorithms.
 Both VisIt and Paraview optionally replace their traditional scan-conversion rendering with ray tracing, utilizing Intel's CPU-based OSPRay and Nvidia's GPU-based Optix.
 
-The second advantage of the ray tracing algorithm is that, in some critical cases, it avoids the necessity of reducing data to low-level geometric primitives.
-Critical to the current work is that the rendering of spheres in a rasterization algorithm requires that each sphere be represented as a set of triangles sufficiently large enough to give the appearance of a smooth sphere.
-When there is a large number of spheres to render, rasterization (which recall considers every triangle) becomes very expensive.
+Another advantage of the ray tracing algorithm is that, in some critical cases, it avoids the necessity of reducing data to low-level geometric primitives.
+Critical to the current work is that the rendering of spheres in a rasterization algorithm requires that each sphere be represented as a sufficiently large set of triangles enough to give the appearance of a smooth sphere.
+When there is a large number of spheres to render, rasterization (which considers every triangle) becomes very expensive.
 Instead, using ray tracing, we can represent each sphere by its center and radius; rather than exhaustively intersect a ray against each triangle of a sphere's tessellated representation, we can determine the distance between the ray and the center and ask whether it is less than the radius.
-This feature is apparent in Paraview; while the traditional rasterization back end required a filter to create spheres of each point, the points can be rendered directly using the OSPRay back end with substantially improved performance.
+This feature is apparent in Paraview; while the traditional rasterization back end required a filter to create spheres of each point, the points can be rendered directly using the OSPRay backend with substantially improved performance.
 
@@
 When visualizing on multiple nodes in parallel, the resulting image may contain imagery computed on any participating node.
-To get a correct image, visualization systems generally use one of two algorithms.
-First, they can redistribute the primitives to provide all the information necessary to render a screen patch (possibly the entire image).
-Alternatively, each node renders its portion of the data independently and then uses a distributed-parallel depth-based compositing algorithm to reconcile these partial results into a final image.
-Each has advantages: if the amount of data required to render the image is small, it can be gathered cheaply and then rendered interactively with no further data movement.
+To get a correct image, visualization systems generally use one of two algorithms, described in Molnar et.
+al ~\cite{Molnar94} as \textit{sort-first\textit} and \textit{sort-last} rendering.
+The sort-first algorithm redistributes the data to provide all the information necessary to render a screen patch (possibly the entire image).
+In the sort-last algorithm, each node renders its portion of the data independently and then uses a distributed-parallel depth-based compositing algorithm to reconcile these partial results into a final image.
+Each has its advantages: if the amount of data required to render the image is small, it can be gathered cheaply and then rendered interactively with no further data movement.
 On the other hand, when the data is large, the gather is expensive (and potentially exceeds the memory available on the rendering node(s)), and the post-rendering compositing step is preferable.
 
-When advanced global lighting is required, we note that the second of these algorithms will not work.
+When advanced global lighting is required, we note that sort-last rendering will not work.
 Without shadows, the correct shading for a surface is determined independently of other surfaces and thus can be performed locally before compositing.
-When shadows are required, the surface can be shadowed by a surface \textit{that is not present on current node}.
+However, when shadows are required, it is possible that a surface can be shadowed by another surface \textit{that is not present on current node}.
 Thus post-rendering compositing cannot create a correct image, leaving gathering the data as the alternative.
 
-Galaxy, developed at TACC, differs from OSPRay and Optix in that it is based on a regular partitioning of the computational space, with all the data contained in each partition local to the node responsible for the partition.
-This assumption - derived from the spatial partitioning commonly used in distributed-memory simulation systems - allows Galaxy to do \textit{most} of the work locally, considering (on average) only 1/N of the data, and transferring rays in a nearest-neighbor pattern only when rays (primary and secondary) exit one partition and enter another.
-Rays are retired when they are resolved, and results are transferred to a potentially remote frame buffer.
+Galaxy, developed at TACC, instead relies on \textit{spatial partitioning of the data} to implement Molnar's \textit{sort-middle} algorithm.
+Spatial partitioning - generally a natural data decomposition in HPC simulation - ensures that all the data that may be relevant in region of space is resident on the process responsible for that region.
+Galaxy's cooperating processes can therefore trace rays \textit{within their partition space} independently of data in other spaces.
+Within the partition, rays are either retired (by reaching full opacity in volume rendering, or by encountering surfaces), or they reach the partition boundary.
+When a boundary is encountered, the rays are transmitted to the process responsible for the partition across the boundary.
+Thus the lighting of a surface hit point within a partition can be resolved by a secondary ray that first determines whether the light is occluded by local data, and if not, by data resident in other processes on the way to the light source.
+
+Note that Galaxy is actually a framework for the management and exchange of rays between processes; it does not actually perform ray/surface or ray/volume intersections itself.
+Rather, it relies on OSPRay (i.e., OSPRay's Embree and OpenVKL underpinnings) to perform the \textit{local} ray tracing.
+This could equally well be performed by Optix.
+For more information on Galaxy, please see Abram et.
+al.
+\cite{abram2018galaxy}.
 
@@
 Of course, in-situ visualization largely precludes exploratory visualization: the in-situ analysis will only produce the anticipated results at set-up time.
-With post-processing, the entire dataset is available when the user sits down in front of the visualization system; the user can adjust viewpoints and parameters interactively during his session, a powerful capability lost in in-situ visualization.
+With post-processing, the entire dataset is available when the user sits down in front of the visualization system; the user can adjust viewpoints and parameters interactively during their session, a powerful capability lost in in-situ visualization.
 However, Ahrens et al.
@@
 Therefore, we can set up an in-situ process that generates as many images of the data - varying viewpoints and parameters - as the user anticipates and wants.
+Offering in-situ visualization capabilities incurs additional compute cost at run-time.
+However, post-processing visualization also has this addition compute cost, which is likely to be higher than in-situ visualization due to increased storage and latency requirements.
 Ahrens et al.
@@
 \section{Study Detail}
-The 3D Oso landslide region of interest spans 1500 x 1500 x 270 m.
-This domain is discretized into 4.2 million material points and a background grid of 8 x 8 x 8 m.
-In most regions, we use two material points in each direction (8 material points per cell), while the sliding area has a higher resolution with four material points in each direction (64 points per cell).
-A Mohr-Coulomb material model with a Total Stress Approach is used to simulate the undrained runout evolution of the Oso landslide in MPM.
+The 3D Oso landslide region of interest spans $1500 \times 1500 \times 270$ m.
+This domain is discretized into 4.2 million material points and a background grid with a cell size of $8 \times 8 \times 8$ m.
+In most regions, we use two material points in each direction (8 material points per cell), while the landslide runout region has a higher resolution with four material points in each direction (64 points per cell).
+A Mohr-Coulomb material model with a Total Stress strength parameters is used to simulate the undrained runout evolution of the Oso landslide in MPM.
 
 In this work, we consider in-situ visualization for MPM.
-In particular, we note that the data for visualization is simple: a large set of points with scalar attributes (currently ~4.2M and potentially billions), scattered among a large number of parallel processes.
+In particular, we note that the data for visualization is simple: a large set of points (currently ~4.2M and potentially billions) with scalar attributes, scattered among a large number of parallel processes.
 The objective is to generate potentially many images at an appropriate temporal frequency and minimize the need for MPM to dump timestep data.
@@
 
-We chose Galaxy, our ray-tracing-based visualization system, to interface to MPM.
-Galaxy has several advantages that make it attractive in this use; using the compact representation of spheres as the center with radius, rendering a significant number of spheres efficiently, and scaling as necessary to provide the desired performance.
-Ideally, we would like to balance the performance of the MPM simulation with the Galaxy visualization; by developing an M:N point-to-point interface between MPM worker processes and Galaxy processes, we can vary the number of processes in each so that they run at approximately the same pace.
-When out of balance, we can choose to drop frames of the visualization by having MPM continue without pausing if Galaxy is not ready, or not - by having MPM wait for Galaxy to \textit{be} ready.
+Ray tracing is a natural solution for rendering MPM particle data.
+Rather than represent each particle as a set of primitive surface elements (polygons, necessary for traditional scan-conversion rendering), we can represent particle data far more compactly as a center point and radius.
+
+We interface Galaxy, our ray-tracing-based visualization system, and CB-Geo MPM to perform in-situ visualization of the Oso landslide.
+As described above, Galaxy is able to perform global illumination on our data \textit{without gathering all the data to each node}.
+Galaxy has also been shown to scale to over a billion particles and demonstrated to scale to 128 nodes; see the \textit{Cosmic Web} test case in Abram et.
+al.
+\cite{abram2018galaxy}.
+
+Two concerns arose in developing the MPM-Galaxy interface.
+First, we would like to balance the performance of the MPM simulation with the Galaxy visualization by varying the ratio of MPM processes to Galaxy processes.
+Second, the spatial partitioning layout in MPM is not absolute; while each MPM process manages a relatively compact region of space, these regions overlap and do not form the clean Cartesian boundaries necessary for Galaxy.
+To address these problems, we developed an asynchronous M to N data transfer scheme between M MPM processes and N Galaxy processes that incorporates a sorting step.
+This is implemented by communicating an appropriate layout of partitions, given the number of Galaxy processes in play and the bounding box of the MPM computational space, to MPM at data transfer time.
+Each MPM process then traverses its particle data, determines which Galaxy process's partition space the particle lies, and transfers them accordingly.
+The M:N point-to-point interface between M MPM worker processes and N Galaxy processes offers better load balancing performance.
+We can vary the number of processes for MPM and Galaxy so that they run at approximately the same pace.
+When out of balance, we can
+choose to drop frames of the visualization by having MPM
+continue without pausing for Galaxy (if Galaxy is not ready), or by having MPM wait for Galaxy to be ready.
 
@@
 We use a sockets-based interface for this preliminary work, enabling the two components to communicate via the IP protocol.
-While this enables us to position the components anywhere with IP connectivity, by placing the two on the sample supercomputer, we are able to use the full parallel bandwidth of the supercomputer's interconnect to communicate between MPM and Galaxy worker processes.
+While this enables us to position the components anywhere with IP connectivity, by placing the two on the same supercomputer, we are able to use the full parallel bandwidth of the supercomputer's interconnect to communicate between MPM and Galaxy worker processes.
 
@@
 
-Periodically, the MPM root process (at its synchronization point) attempts to connect to this server based on a host/port pair given in its configuration file (we generally leave this as localhost/1900 and use an SSH tunnel to connect that local port to wherever the Galaxy root process is running).
+Periodically, the MPM root process (at its synchronization point) attempts to connect to this server via a socket.
 A small amount of control information is passed when the connection is established, including a list of host/port pairs for listening Galaxy worker processes and the number of MPM worker processes transferring data.
@@
 Once data has arrived on the Galaxy side, it is re-distributed into a Galaxy-compliant partitioning of the computational space.
-As mentioned above, Galaxy relies on a regular subdivision of space - which is not the case for data acquired from MPM, particularly in that multiple MPM workers can send data to each Galaxy node.
+As mentioned above, Galaxy relies on a regular subdivision of space - which is not the case for data acquired from MPM, particularly when multiple MPM workers can send data to each Galaxy node.
 Since each Galaxy worker knows the overall partitioning, this reshuffle is performed using point-to-point messaging via MPI.
@@
 The JSON consists of several sections describing: the data objects, the distinct visualizations, listing the data objects and operators to be applied to the lighting environment, a list of cameras, and renderer parameters.
-Names in the data object section correspond to the variables in MPM; at setup time, Galaxy creates empty distributed containers for each named object.
-At run time, the results of reshuffling data received from MPM replace the contents (if any) of these distributed data objects.
+Names in the data object section correspond to the variables in MPM.
+At setup time, Galaxy creates empty distributed containers for each named object.
+At run time, as the material points traverse the background grid, the reshuffling results from the MPM simulations (if any) will replace the contents of the distributed datasets.
 
@@
 \section{Results}
-\begin{figure}[!tbp]
-    \includegraphics[width=\linewidth]{figs/timestep-0000.png}
-    \caption{Oso Landslide early}
-    \label{fig:oso-early} 
-\end{figure}
-\vspace{0.2cm}
-\begin{figure}[!tbp]
-  \centering
-    \includegraphics[width=\linewidth]{figs/timestep-0253.png}
-    \caption{Oso Landslide late}
-    \label{fig:oso-late}
-\end{figure}
 
-Given a checkpoint restart file at step 50,000 of the MPM Oso simulation, we ran two versions of the next 15,000 steps to evaluate the cost of in-situ visualization on the simulation.
-Using four nodes of Stampede2 at TACC, each running four MPM worker processes, we first ran 15,000 steps without in-situ visualization but saving checkpoint files every 500 steps.
+\begin{figure*}[!tbp]
+    \begin{subfigure}[b]{0.495\textwidth}
+        \includegraphics[width=\textwidth]{figs/frame_1_00000.png}
+        \caption{runout after 50 s}
+        \label{fig:oso-50} 
+    \end{subfigure}
+    \begin{subfigure}[b]{0.495\textwidth}
+        \includegraphics[width=\textwidth]{figs/frame_1_00243.png}
+        \caption{runout after 74 s}
+        \label{fig:oso-74} 
+    \end{subfigure}
+    \caption{In-situ rendering of MPM simulation with Galaxy. The color gradient shows the amount of particle displacements from their original position.}
+    \label{fig:timing}
+\end{figure*}
+
+Given a checkpoint restart file at step 50,000 (actual runout time of 50 s) of the MPM Oso simulation, we ran two versions of the next 25,000 steps to evaluate the cost of in-situ visualization on the simulation.
+Using four nodes of Stampede2 at TACC, each running four MPM worker processes, we first ran 15,000 steps without in-situ visualization but saving checkpoint files every 5,000 steps, resulting in five time step datasets over this interval.
 In this case, the wall-clock time required to run each 5,000 steps - as determined by timestamps on output files - was approximately 2 hours, 23 minutes.
+Figures \ref{fig:oso-50} and \ref{fig:oso-74} visualize the state of this run at the beginning and end of the simulation interval, using a color-scale to show the distance each particle has traveled from its original location.
+Clearly, the simulation has made substantial progress in this time interval.
 
-We then reran the same section of the simulation using the MPM/Galaxy interface to transfer data from MPM to Galaxy every 100 steps and using Galaxy to generate the first two and later three images of each timestep (having noted the runout spread outside our initial camera views) of each step.
-Figures \ref{fig:oso-early} and \ref{fig:oso-late} show two views of this run.
+We then reran the same interval of the simulation with the same checkpoint dumps, but now using the MPM/Galaxy interface to transfer data from MPM to Galaxy every 100 steps (0.1 s of real-time runout, where a difference in the runout can be perceived).
+In-situ visualization with Galaxy/MPM integration offers insights into the dynamics of very fast runouts, such as the Oso landslide (total duration of runout is 120s) with a very fine temporal resolution, which are otherwise infeasible using traditional post-processing visualization.
+We found that Galaxy was substantially faster than MPM, even when running on a single node.
+To make full use of the I/O bandwith available on that node, we use a M:N interface of 4:1 with four MPM worker nodes to one Galaxy worker nodes.
+With this setup, Galaxy required approximately 6.5 seconds for each timestep to receive the data, repartition it, and render the visualizations - far less than the three minutes required by MPM to run through 100 steps.
+In this second case with in-situ visualization, the wall-clock time needed by MPM to run through 5,000 steps was approximately 2 hours, 26 minutes, or 2 percent more than the first, non-visualized case.
+Galaxy renderings of the Oso landslide during the early and late stages of the runout are shown in~\cref{fig:oso3,fig:timing}.
 
-In this case, 
-we ran Galaxy using four worker processes on a single node, so there was a 4:1 ratio of both sending processes to receiving process and sending nodes to receiving nodes.
-Galaxy required approximately 3 seconds for each timestep to receive the data, repartition it, and render the visualizations - far less than the ~three minutes required by MPM to run through 100 steps.
-In this second case, the wall-clock time needed by MPM to run through 5,0000 steps was approximately 2 hours, 26 minutes, or 2 percent more than the first, non-visualized case.
+Galaxy supports including additional views at runtime as we observe interesting dynamics for visualization as the simulation progresses.
+Galaxy initially generated two images per timestep and later three, when we observed the runout spreading outside our initial camera views.
 
@@
 
-The above demonstrates that, at this scale, the time cost of in-situ transfer of data from this simulation to a visualization application located on a separate partition of the same supercomputer is negligible.
-Further, the time required to generate the three simple views of the data for every 100 timesteps is much less than the cost of simulating 100 timesteps, indicating that more elaborate visualization (for example, generating many views of each timestep, enabling Cinema-based, real-time interaction the visualization) is achievable without impacting the simulation.
+In our study with 4.7 million material points, the time cost of in-situ data transfer from simulation to a visualization application located on a separate partition of the same supercomputer is negligible.
+A many-to-many communications strategy using point-to-point communications enables us to use the full cross-sectional internal bandwidth of the supercomputer interconnect to move data from a simulation to an external concurrently running analysis component with minimal impact on the simulation.
+Further, the time required to generate the three simple views of the data for every 100 timesteps is much less than the cost of simulating 100 timesteps - though clearly this depends on both the simulation and the visualization.
+This leads to several options.
+We could reduce the resources allocated to visualization for better balance of \textit{this} workflow (though not in this case, since we use a single node for visualization).
+Alternatively, we could increase the work necessary on the visualization side, perhaps by generating \textit{many} views of each timestep, enabling Cinema-based, real-time interaction with visualization \cite{Ahrens2014cinema}.
+Finally, we could increase the frequency of timesteps.
+In any event, our architecture enables these factors to be considered independently, by varying the allocation of resources to the in-situ analysis.
 
-We realized that a new view was necessary to capture the full extent of the run-out at mid-run.   We were able to re-parameterize the visualization on the fly to add this additional view.
-
-\subsection{Future work}
-
-Clearly, the test example we have worked with - the Oso landslide at a relatively coarse resolution does not tax our system.
+Clearly, the test example we have worked with - the Oso landslide at a relatively coarse resolution, does not tax our system.
 However, as observed above, regional-scale simulation of landslides resulting in kilometer-scale runout will require billions of particles rather than millions.
-We believe our approach - notably using a scalable visualization component (Galaxy, \cite{abram2018galaxy}) and a scalable N:M interface architecture enables us to re-scale the visualization process as necessary to achieve in-situ visualization of simulation data with minimal impact on the simulation process.
+We believe our approach - notably using a scalable distributed visualization component (Galaxy, \cite{abram2018galaxy}) and a many-to-many, point-to-point, in-situ interface that leverages the parallel internal bandwidth of the supercomputer, provides a means for effectively visualizing full-scale simulations.
 
@@
 % if you will not have a photo at all:
-% \begin{IEEEbiographynophoto}{John Doe}
-% Biography text here.
-% \end{IEEEbiographynophoto}
-
-%\vfill
+\vfill\null
+%\columnbreak
+\begin{IEEEbiographynophoto}{Greg Abram}
+is a visualization researcher in the at the Texas Advanced Computing Center, University of Texas at Austin, USA, 78758.
+His research interests focus on parallel visualization systems.
+Dr.
+Abram received a Ph.D.
+in Computer Science from the University of North Carolina at Chapel Hill.
+Contact him at gda@tacc.utexas.edu.
+\end{IEEEbiographynophoto}
+\begin{IEEEbiographynophoto}{Andrew Jay Solis}
+is a visualization researcher in the at the Texas Advanced Computing Center, University of Texas at Austin, USA, 78758.
+His research interests include extended realities, HCI, scientific computation, and parallel computing.
+Solis received a B.S.
+in Computer Science from the University of Texas at Austin.
+Contact them at asolis@tacc.utexas.edu.
+\end{IEEEbiographynophoto}
+\begin{IEEEbiographynophoto}{Yong Liang}
+is a postdoctoral researcher in the Civil and Environmental Engineering, University of California at Berkeley, USA, 94720.
+His research interests include fracture modeling and material point method.
+Yong received his Ph.D.
+from Tsinghua University, China.
+Contact them at yliang\_sn@berkeley.edu.
+\end{IEEEbiographynophoto}
+\begin{IEEEbiographynophoto}{Krishna Kumar}
+is an Assistant Professor in the Civil, Architectural and Environmental Engineering, University of Texas at Austin, USA, 78751.
+His research interests include high-performance computing, numerical modeling, and explainable-AI.
+Krishna received his Ph.D.
+in Engineering from the University of Cambridge, UK.
+Contact them at krishnak@utexas.edu.
+\end{IEEEbiographynophoto}
+\vfill
 
